EDA on Loan dataset from Prosper
========================================================

This is an exploration of loan dataset that was downloaded from Prosper. Prosper
is a marketplace lending platform where individuals can invest or borrow money. 
It allows people to invest $25 or more fr a given loan listing.


```{r echo=FALSE, message=FALSE, warning=FALSE, packages}

# Load all of the packages used in the analysis

# install.packages('GGally')
# install.packages('lattice')
# install.packages('MASS')
# install.packages('reshape2')
# install.packages('dplyr')
# install.packages('RCurl')
# install.packages('tidyr')
# install.packages('gridExtra') 

library(gridExtra) 
library(reshape2)
library(tidyr)
library(dplyr)
library(RCurl)
library(GGally)
library(lattice)
library(MASS)
library(RColorBrewer)
library(ggplot2)

```


**Dimensions of the loan dataset:**
```{r echo=FALSE, message=FALSE, warning=FALSE, Load_the_Data}

# Load the Data
setwd('C:/Users/pneupane/Documents/study topics/udacity -Data Analyst ND/P4 EDA/project')
loan = read.csv("prosperLoanData.csv")
dim(loan)

```

The loan dataset has 113,937 loans along with 81 variables for each loan. The 
variables specify the attributes of the loan and well as the individual requesting
the loan. 


------

# Univariate Plots Section

**Here is a quick list of variables in the dataset**

```{r echo=FALSE, Univariate_Plots}

str(loan)

```


**Here is a quick summary of the dataset for selected variables of interest:**
```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots1}

summary (loan[c("LoanStatus", "Term","BorrowerAPR","EstimatedLoss","EstimatedReturn",
                "CreditScoreRangeUpper","CurrentCreditLines","OpenCreditLines",
                "ProsperScore", "IncomeRange",
                "TotalCreditLinespast7years",
                "TotalInquiries","DelinquenciesLast7Years",
                "DebtToIncomeRatio","ProsperPrincipalBorrowed"
                )])

```

We'd want to check LoanStatus to see if the loan was completed, or if it was 
considered Bad loan. Count of Loan Status are shown in actual and log10 scale.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots2,fig.height=3,fig.width=8}

ggplot(aes(x=LoanStatus), data = loan) + 
  geom_histogram(stat="count") +
  coord_flip() +
  labs(x="Loan Status",y="Count of loans")


ggplot(aes(x=LoanStatus), data = loan) + 
  geom_histogram(stat="count") +
  coord_flip() +
  labs(x="Loan Status",y="Count of loans (log scale)") +
  scale_y_log10()

```

To make it easier to visualize good loans from bad loans, we convert all loans 
that are past due for more than 30 days, Defaulted and Chargedoff loans as 
**'Bad loans'**. Count of Loan Status are shown in actual and log10 scale.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots3,fig.height=3,fig.width=8}


### Combine Past Due (>60 days), Defaulted and Chargedoff into one bucket
#table(loan$LoanStatus)
loan$LoanStatus_mod = as.character(loan$LoanStatus)
loan$LoanStatus_mod[loan$LoanStatus %in% 
                      c("Past Due (31-60 days)","Past Due (61-90 days)", 
                        "Past Due (91-120 days)", "Past Due (>120 days)",
                        "Chargedoff", "Defaulted")] ="Bad Loan"

loan$LoanStatus_mod[loan$LoanStatus %in% 
                      c("Past Due (1-15 days)",  
                        "Past Due (16-30 days)")] ="Past Due (<30 days)"

loan$LoanStatus_mod[loan$LoanStatus 
                    %in% c("FinalPaymentInProgress")] ="Completed"


### Remove cancelled
loan = loan[loan$LoanStatus != "Cancelled",]
#table(loan$LoanStatus_mod)


#print(levels(loan$IncomeRange))

### reorder levels for Income Range
loan$IncomeRange = factor(loan$IncomeRange,levels=
                            c("Not displayed","Not employed","$0","$1-24,999",     
                              "$25,000-49,999", "$50,000-74,999", "$75,000-99,999",
                              "$100,000+")
)
  
# Display histogram                       
ggplot(aes(x=LoanStatus_mod), data = loan) + 
  geom_histogram(stat="count", binwidth = .5) +
  coord_flip() +
  labs(x="Loan Status Modified",y="Count of loans")


# Display histogram                       
ggplot(aes(x=LoanStatus_mod), data = loan) + 
  geom_histogram(stat="count", binwidth = .5) +
  coord_flip() +
  labs(x="Loan Status Modified",y="Count of loans (log scale)") +
  scale_y_log10()

```



**Explore Prosper Score.** This score is built using Prosper historical data.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots4, fig.height=3,fig.width=8 }

###---------------------------------------------------------------------
#### Histogram of Prosper score
ggplot(aes(x=ProsperScore), data = loan) + 
  geom_histogram(stat="count") +
  coord_flip() +
  labs(x="Prosper Score",y="Count of loans")

ggplot(aes(x=ProsperScore), data = loan) + 
  geom_histogram(stat="count") +
  coord_flip() +
  labs(x="Prosper Score",y="Count of loans (log scale)") +
  scale_y_log10()

```


ProsperScore is supposed to have values ranging from 1-10. But it also seem to 
have 11. 
```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots5, fig.height=3,fig.width=8 }

table(loan$ProsperScore)

```

**Explore Term values:**

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots6 }

###---------------------------------------------------------------------
#### Freq counts  of Term

a=as.data.frame(table(loan$Term))
names(a)[1] = 'Loan Term'
names(a)[2] = 'Freq count of Loan Term'
print(a)

  
```

There are only 3 terms 12, 36, and 60, and the most common seems to be 36.


**Explore Income Range:**

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots7, fig.height=3,fig.width=8 }

###---------------------------------------------------------------------
#### Histogram of Income Range
ggplot(aes(x=IncomeRange), data = loan) + 
  geom_histogram(stat="count") +
  labs(x="Income Range",y="Count of loans")


ggplot(aes(x=IncomeRange), data = loan) + 
  geom_histogram(stat="count") +
  labs(x="Income Range",y="Count of loans (log scale)")+
  scale_y_log10()

```

Most frequent Income Range is $25k - 75k,


**Explore Total number of Inquiries in past 6 months:**

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots7a, fig.height=3,fig.width=8 }

###---------------------------------------------------------------------
#### Histogram of Total Inquiries when the credit was pulled

ggplot(aes(x=InquiriesLast6Months), data =  loan) + 
  geom_histogram(binwidth=.5)

# Boxplot
ggplot(aes(x=1,y=InquiriesLast6Months), data = subset(loan,InquiriesLast6Months>0)) + 
  geom_boxplot()  +
  labs(y="Inquiries last 6 months") +
  coord_flip()

```

There seems to be outliers that are large distance away from the IQR range of data. 
We try to limit the upper range at 30.

Check how frequency for each InquiriesLast6Months value:
```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots7b, fig.height=3,fig.width=8 }

table(loan$InquiriesLast6Months)

```

Majority of records have anywhere between 0-2 inquiries in last 6 months. To make 
those more visible, lets focus on only records with 0-4 inquiries in last 6 months.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots7c, fig.height=3,fig.width=8 }

### 
ggplot(aes(x=InquiriesLast6Months), data =  subset(loan,InquiriesLast6Months<5)) + 
  geom_histogram(binwidth=.5) +
  scale_x_continuous(breaks=seq(0,20,1)) +
  labs(x="Inquiries last 6 months",y="Count of loans") +
  coord_cartesian(xlim=c(0,4))

ggplot(aes(x=1,y=InquiriesLast6Months), data = subset(loan,InquiriesLast6Months<5)) + 
  geom_boxplot()  +
  labs(y="Inquiries last 6 months")+
  coord_flip()

```
  
Most inquiries in last 6 months tend to between 0 and 2.



**Explore Debt to Income Ratio:**

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots8, fig.height=3,fig.width=8 }

###---------------------------------------------------------------------
#### Histogram of Debt to Income Ratio
ggplot(aes(x=DebtToIncomeRatio), data = loan) + 
  geom_histogram(binwidth = .01) +
  scale_x_continuous(breaks=seq(0,1,.05)) +
  labs(x="Debt to Income Ratio between 0 and 1",y="Count of loans") +
  coord_cartesian(xlim=c(0,1))

# Boxplot
ggplot(aes(x=1,y=DebtToIncomeRatio), data = loan) + 
  geom_boxplot() +
  scale_y_continuous(limits=c(0,1), breaks=seq(0,1,.05)) +
  labs(y="Debt to Income Ratio between 0 and 1") +
  coord_flip() 

```

Debt to Income Ratio can range from 0 to 10.0, and IQR range is very narrow 
(between ~0.2 to ~0.3). 



**Explore Credit Scores (Lower Range):**

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots9, fig.height=3,fig.width=8 }

###---------------------------------------------------------------------
#### Histogram of Credit Score Lower value
ggplot(aes(x=CreditScoreRangeLower), data = loan ) + 
  geom_histogram(stat="count")  +
  labs(x="Credit Score (Lower Range)>0",y="Count of loans")

# Boxplot
ggplot(aes(x=1,y=CreditScoreRangeLower), data = loan) + 
  geom_boxplot() +
  labs(y="Credit Score(Lower Range)") +
  coord_flip() 

```

There seems to be loans without or very low credit scores, Credit Score. So we
only look at those with credit scores larger than 500.

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots9a, fig.height=3,fig.width=8 }

###---------------------------------------------------------------------
#### Histogram of Credit Score Lower value
ggplot(aes(x=CreditScoreRangeLower), data = subset(loan,CreditScoreRangeLower>0)) + 
  geom_histogram(stat="count") +
  scale_x_continuous(breaks=seq(500,900,50)) +
  labs(x="Credit Score (Lower Range)>0",y="Count of loans")

# Boxplot
ggplot(aes(x=1,y=CreditScoreRangeLower), data = subset(loan,CreditScoreRangeLower>0) ) + 
  geom_boxplot() +
  scale_y_continuous(limits=c(500,900), breaks=seq(500,900,50)) +
  labs(y="Credit Score(Lower Range)") +
  coord_flip() 

```

Most Credit Scores fall between 650 and 750.


------

# Univariate Analysis


### What is the structure of your dataset?

There are 113937 observations and 81 variables. Only 15 variables out of the 
original 81 are considered of interest for this analysis.

* Term: There are 3 values for Terms (12,36, and 60). 36 month seems to be most 
common.

* LoanStatus_mod: Has values of Bad Loan, Completed, Current and 
Past Due (<30 days). Cancelled loans are removed.


* ProsperScore: Ranges from 1-11. Although, according to the documentation, it 
should only be 1-10.

* CreditScoreRangeLower: Ranges from about 500 to 900. Majority of individuals 
have credit score between 650 and 750. Median value is 699.

* IncomeRange: Values include 'Not displayed', 'Not employed', '$0', ranges of 
values from $1 to $100,000, with increments of $25,000, and anything equal to or
larger than $100,000. Most income seem to be in the ranges '$25,000-49,999' and
$50,000-74,999.

* DebtToIncomeRatio: IQR range is 0.14 - 0.32. There are some ratios higher than 
1, but those are considered outliers.

* TotalInquiries: For most cases, most inquiries seem to be between 0 to 6. Mean 
number of inquiries is 5.58.



### What is/are the main feature(s) of interest in your dataset?

The main features of interest are as follows: 
"LoanStatus", "Term","BorrowerAPR", "CreditScoreRangeLower", "ProsperScore", 
"IncomeRange", "TotalInquiries","DelinquenciesLast7Years", "DebtToIncomeRatio"

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

Other features that might help are:
"BorrowerAPR", "ProsperRating..Alpha.", "EmploymentStatus", 
"EmploymentStatusDuration", "IsBorrowerHomeowner", etc.

### Did you create any new variables from existing variables in the dataset?

Yes. I created LoanStatus_mod with consolidated bad-loan values. 

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

Removed LoanStatus="Cancelled". The objective of this analysis is to study what
differentiates good loan from bad loan. If the loan was cancelled, it did not 
contribute to the objective.


Reordered the levels for IncomeRange to make them appear in the plots in desired
order.


------

# Bivariate Plots Section

**Study the relationship between Loan Status and Prosper Score:**

Histograms of Prospect Scores by Loan Status
```{r echo=FALSE,  message=FALSE, warning=FALSE, Bivariate_Plots0, fig.height=3,fig.width=8 }

### Relationshop between Loan Status and ProsperScore
#### Is ProsperScore a good indicator that the loan will be paid off??

ggplot(aes(x=ProsperScore), data = loan) + 
  geom_histogram(stat="count") +
  coord_flip() +
  labs(x="Prosper Score",y="Count of loans") +
  facet_grid(.~LoanStatus_mod)

# ------ Loans with higher Prosper-score seems to have smaller proportion of bad loans
```

Bad Loan have Prosper Score concentrated towards the middle of the score range.
Whereas Completed loans are concentrated at the higher range of 
Prosper Score.


Next is proportions of Loan Statuses for each Prosper Score bucket.
```{r echo=FALSE,  message=FALSE, warning=FALSE, Bivariate_Plots1, fig.height=3,fig.width=8 }

# Remove Current Loan Status
ggplot(aes(x=ProsperScore), data = loan) +
  geom_bar(stat="count",aes(fill=LoanStatus_mod), position="fill") +
  labs(y="Percentage of counts") +
  scale_x_continuous(breaks=seq(0,10,1))
# ------ Ditto. Prosper Scores below 7 seem to have same proportion of bad loans

```

Loans with higher Prosper-score seems to have smaller proportion of bad loans, 
compared to those with lower scores.



**Explore relationship between Loan Status and Credit-Score:**
```{r echo=FALSE,  message=FALSE, warning=FALSE, Bivariate_Plots2, fig.height=3,fig.width=8 }
###---------------------------------------------------------------------
### Relationship between Loan Status and Credit-Score
ggplot(aes(x=LoanStatus_mod ,y=CreditScoreRangeLower), data = subset(loan,CreditScoreRangeLower>0)) +
  geom_boxplot() +
  labs(x="Loan Status Modified", y="Credit Score Range (Lower)") 

# ------ IQR range for Bad Loans are lower compared to other Loan Status
```

IQR range of Credit Scores for Bad Loans are lower compared to that of other Loan 
Statuses.


**Explore relationship between Loan Status and Current Delinquencies:**
```{r echo=FALSE,  message=FALSE, warning=FALSE, Bivariate_Plots3, fig.height=3,fig.width=8 }
###---------------------------------------------------------------------
### Relationships between Number of Current Delinquencies for each Loan Status
ggplot(aes(x=LoanStatus_mod ,y=DelinquenciesLast7Years), 
       data = subset(loan, DelinquenciesLast7Years>0)) +
  geom_boxplot() +
  coord_cartesian(ylim=c(0,20))
```

Number of Delinquencies in past 7 years does not seem to determine if the loan 
will be 'bad'. 


**Explore relationship between Inquiries in last 6 months  and Debt to income ratio:**
```{r echo=FALSE,  message=FALSE, warning=FALSE, Bivariate_Plots4, fig.height=3,fig.width=8 }

###---------------------------------------------------------------------
### Relationships between Inquiries in last 6 months  and Debt to income ratio
ggplot(aes(x=DebtToIncomeRatio,y=InquiriesLast6Months), data = loan) +
  geom_jitter(alpha=1/20) + 
  coord_cartesian(xlim=c(0,1),ylim=c(0,50))

```

Overall, there doesn't seem to be any relationship between Inquiries in last 
6 months  and Debt to income ratio


**Explore relationship between Income Range and Loan Status:**
```{r echo=FALSE,  message=FALSE, warning=FALSE, Bivariate_Plots5, fig.height=3,fig.width=8 }

###---------------------------------------------------------------------
### Relationshop between Income Range and Loan Status
ggplot(aes(x=IncomeRange), data = loan) +
  geom_bar(stat="count",aes(fill=LoanStatus_mod),position="fill") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

``` 

Loans with higher Income Range have smaller rate of Bad-Loans. 

Next, we study the same relationship but without 'Current' loan status.
```{r echo=FALSE,  message=FALSE, warning=FALSE, Bivariate_Plots5a, fig.height=3,fig.width=8 }

###---------------------------------------------------------------------
### Relationshop between Income Range and Loan Status
ggplot(aes(x=IncomeRange), data = subset(loan,LoanStatus !='Current') ) +
  geom_bar(stat="count",aes(fill=LoanStatus_mod),position="fill") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


``` 

After excluding 'current' loan status, for income range '$100,000+', about 25% 
are classified as Bad Loan. Comparing that to income range '$1-$25,000', only 
about 35% of loans are classified as Bad loans.


**Explore relationship between Income Range and Debt to Income Ratio:**
```{r echo=FALSE,  message=FALSE, warning=FALSE, Bivariate_Plots6, fig.height=3,fig.width=8 }

# Income Range and Debt to Income Ratio
ggplot(aes(x=IncomeRange ,y=DebtToIncomeRatio),data=loan ) +
  geom_boxplot()

```

Most Debt to income ratio values are concentrated <1. So we zoom-in:

```{r echo=FALSE,  message=FALSE, warning=FALSE, Bivariate_Plots6a, fig.height=3,fig.width=8 }

# Income Range and Debt to Income Ratio
ggplot(aes(x=IncomeRange ,y=DebtToIncomeRatio),data=loan ) +
  geom_boxplot() +
  coord_cartesian(ylim=c(0,1))


```

Loans with low income have debt to income ratio more spread out (with some exceeding 10.0). 
Whereas, Loans with higher income consistently have very low debt to income ratio.


**Explore relationships between LoanStatus and Debt to Income Ratio:**

```{r echo=FALSE,  message=FALSE, warning=FALSE, Bivariate_Plots6b, fig.height=3,fig.width=8 }

# Income Range and Debt to Income Ratio
ggplot(aes(x=LoanStatus_mod ,y=DebtToIncomeRatio),data=loan ) +
  geom_boxplot() +
  coord_cartesian(ylim=c(0,1))


```


Debt to Income ratio does not vary a lot for different values of Loan Status. This 
is surprising because I expected the Bad Loans to have higher Debt to Income ratio.


**Pairwise graphs of variables among variables that are not of main interest:**
```{r echo=FALSE,  message=FALSE, warning=FALSE, Bivariate_Plots7, fig.height=6,fig.width=6 }

ggpairs(data=  loan, 
        columns=c("BorrowerAPR","EstimatedReturn","EstimatedLoss", "AvailableBankcardCredit"),
        axisLabels= "show")

```

Among variables that are not of Interest, BorrowerAPR vs. Estimated Return, and 
Estimated Return vs. EstimatedLoss have highest correlation. I guess it makes sense
that loans with higher APR have higher Estimated Returns. I expected Borrower APR
to have high negative relation with EstimatedLoss, but that doesn't seem to be the case.


------

# Bivariate Analysis


### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

Prosper Score for bad loans are concentrated more towards middle of the score range.

Credit score range for Bad loans (600-700) is lower than compared to completed loans
(650-750).

Number of Delinquencies in tha pst 7 years do not seem to determine the loan
status.

As expected, loans with higher Income Range have higher proporiton of bad loans 
compared to that of completed loans.




### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

Borrower APR has positive relationship with Estimated Return, and with Estimated
Loss.

There are some Listing Categories that have higher Estimated  




### What was the strongest relationship you found?

The strongest relationship is that of Borrower APR against Estimated Return 
and Estimated Loss.



------

# Multivariate Plots Section


**Explore relationship between ProsperScore and Loan Status, grouped by credit score 
ranges:**
```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots1, fig.height=3,fig.width=8}


# Credit Score Buckets
levels = c(-Inf,500,600,700,800,900,Inf)
labels = c("<500","500-600","600-700","700-800","800-900","900+")

loan = 
  loan %>%
    mutate(CreditScoreBucket=cut(CreditScoreRangeLower,levels, labels=labels))


# Split by Credit Score
ggplot(aes(x=ProsperScore), data = subset(loan,LoanStatus !='Current' & !(CreditScoreBucket %in% c("<500","900+")) & !is.na(CreditScoreBucket) ) ) +
  geom_bar(stat="count", aes(fill=LoanStatus_mod), position="fill") +
  labs(y="Percentage of counts") +
  scale_x_continuous(breaks=seq(0,10,1)) +
  facet_grid(.~ CreditScoreBucket )


```

Credit scores 800 to 900 seem to have very less proportion of Bad loans for higher Prosper Scores.
But for lower Prosper scores in the same credit range, the proportion of Bad loans are 
comparable to other Credit Score ranges. This tells us that just looking at
the Credit Score might not be enough to determine if the loan will be good or bad.


```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots1a, fig.height=8,fig.width=8}


# Split by Credit Score
ggplot(aes(x=ProsperScore), data = subset(loan,LoanStatus !='Current' & !(CreditScoreBucket %in% c("<500","900+")) & !is.na(CreditScoreBucket) & IncomeRange != "Not displayed") ) +
  geom_bar(stat="count", aes(fill=LoanStatus_mod), position="fill") +
  labs(y="Percentage of counts") +
  scale_x_continuous(breaks=seq(0,10,1)) +
  facet_grid(IncomeRange~. )


```

Higher Income ranges and higher Prosper Scores have less proportion of Bad Loans,
compared to lower Income Ranges and Lower Prosper Scores.


**Explore relationship between Credit Score and mean Income Ranges, for each 
Loan Status type:**
```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots2, fig.height=3,fig.width=8}

ggplot(aes(x=IncomeRange,y= CreditScoreRangeLower, fill=LoanStatus_mod  ), data = loan) +
  geom_boxplot(size=1) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + ylim(400, 800)+
  labs(x="Income Range")



```

In general, credit scores for people with bad loans are lower compared to 
that for other groups. Surprisingly, overall credit scores for unemployed people 
is fairly high. For them, mean credit score do not seem to be good indicator
their Loan Status.

Also, for employed people, credit score IQR range for bad loans and 
"Past Due (<30 Days)" seems to be lower than completed and current loans. This 
might indicate that for higher income-range group, it is more likey for people 
with lower credit scored to default on a loan.

So, for employed people, Income range and mean Credit Score seems to be good 
indicator of Loan Status. 


**Explore relationships of Credit Scores, Income Range, and InquiriesLast6Months 
to Loan Status:**

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots2a, fig.height=15,fig.width=8}

ggplot(aes(y=InquiriesLast6Months,x=CreditScoreBucket, 
           fill=LoanStatus_mod), 
       data = subset(loan,!(CreditScoreBucket %in% c("<500","900+")) &
                       !is.na(CreditScoreBucket))
      ) +
  geom_boxplot() +
  facet_grid(IncomeRange~.) +
  labs(x="Credit Score Range bucket",y="Mean Inquiries Last 6 Months") +
  coord_cartesian(ylim=c(0,10))
  
```  

We can see that Bad Loans have slightly higher number of Inquiries in Last 6 Months
for all Credit score range buckets and Income Ranges. Bad Loans in lower credit 
score range (500-600) and those in high range (800-900) seem to have the most
variation in terms of number of Inquiries in Last 6 Months. 

For lower Credit Score Range (500-600), completed loans have more Inquiries in 
Last 6 Months compared to Current loans. This might be because of the fact that 
completed loans includes past loans where investors were more willing to take 
larger risks (and therefore reaped larger rewards). But the investors for current 
loans are less willing to take risks (especially for people with lower credit scores). 
Therefore, we see them investing only on those loans with lower number of Inquiries.


**Study relationship between Income Range and Loan Status, by Income-Verifiable:**

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots3, fig.height=3,fig.width=8}

# Split for verifiable vs. non-verifieable income
ggplot(aes(x=IncomeRange), data = subset(loan,LoanStatus !='Current') ) +
  geom_bar(stat="count",aes(fill=LoanStatus_mod),position = "fill") +
  labs(x = "Income Range", y="Percentage counts") +
  facet_grid(.~IncomeVerifiable) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Is Income Verifiable?")  +
  theme(plot.title = element_text(hjust = 0.5))

```

Non-verifiable Income group seems to have equal percentage of Bad Loans for all 
income ranges. Therefore, for Non-verifiable group, Income Range does not
seem to be good inidcator of Loan Status.


**Explore relationship between Income Ranges and Debt to income Ratio, for each
Loan Status type:**
```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots4, fig.height=3,fig.width=8}

ggplot(aes(x=IncomeRange,y=DebtToIncomeRatio,fill=LoanStatus_mod), data = loan) +
  geom_boxplot( ) +
  scale_colour_brewer(type = 'div')+
  coord_cartesian(ylim=c(0,1)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```


For each Income Range, Bad Loans have slightly lower Debt to Income ratio compared
to Current loans, and slightly higher Debt to Income ratio compared to Completed 
loans. This is counterintuitive because one would expect the Bad Loans to have
higher Debt to Income ratio than any other categories. 



**Explore relationship between Debt to income Ratio and Inquiries in last 6 months, 
for each Loan Status type:**
```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots5, fig.height=3,fig.width=8}

# Mean inquiries in past 6 months for different Loan Types
ggplot(aes(x=DebtToIncomeRatio ,y=InquiriesLast6Months), data = loan) +
  geom_smooth(aes(color=LoanStatus_mod),stat='summary', fun.y=mean) +
  scale_colour_brewer(type = 'div') +
  labs(x="Debt to Income Ratio", y="Inquiries in Last 6 Months")
```

Looking at the mean Inquiries in Last 6 Months for different levels of Debt to 
Income Ratio, it is hard to read the chart because there is too much fluctuation
over a short range. 

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots5a, fig.height=3,fig.width=8}

ggplot(aes(x=.5*round(DebtToIncomeRatio/.5,1) ,y=InquiriesLast6Months), data = loan) +
  geom_smooth(aes(color=LoanStatus_mod),stat='summary', fun.y=mean) +
  scale_colour_brewer(type = 'div') +
  labs(x="Debt to Income Ratio (smoothed)", y="Inquiries in Last 6 Months (smoothed)")

```

So, we smooth the chart by grouping Debt to Income ratio into increments of .5, 
and taking the mean of InquiriesLast6Months. The smoothed plot shows that Bad Loans
have higher Inquiries in past 6 months compared to any other Loan Types.

In the lower range of DebtToIncomeRatio (0-2.5), it is interesting to see than Bad Loans 
deviate from Good Loans by a huge margin in terms of Inquiries in Last 6 Months. 
For DebtToIncomeRatio above 2.5, Loan Status values do not seem deviate a lot from
each other.

Another interesting observation is that alomst all Current loans have DebtToIncomeRatio
below 1, peaking around .7. What can be the reason for this? Maybe this the threshold where 
risk vs reward is optimal?


Next, we attempt to facet the graph by Income Range:

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots5b, fig.height=8,fig.width=8}

ggplot(aes(x=.5*round(DebtToIncomeRatio/.5,1) ,y=InquiriesLast6Months), data = subset(loan, IncomeRange !="$0")) +
  geom_smooth(aes(color=LoanStatus_mod),stat='summary', fun.y=mean) +
  scale_colour_brewer(type = 'div') +
  labs(x="Debt to Income Ratio (smoothed)", y="Inquiries in Last 6 Months (smoothed)") + 
  scale_x_continuous(breaks=seq(0,10,1))+
  facet_grid(IncomeRange~.)

```

In general # of Inquiries are higher for Bad Loans compared to other Loan Statuses,
for all income Ranges and Debt to Income ratio ranges.

For loans with Income Range of $1k to $25k, Not employed and Not displayed, # of 
inquiries are significantly higher for Bad Loans compared to Completed Loans. There
is a huge peak at 1.5.

For Income range $25k or higher, # of inquiries are similar for all loan statuses.


------

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

Loan Status (whether a loan will be Defaulted or not) seems to tbe highly related
to Prosper Score, Credit Score, Income Ranges and Inquiries within last 6 months.

Income Range and Credit Score strengthened each other. Debt to Income Ratio and 
Inquiries within last 6 months also strengthened each other.



### Were there any interesting or surprising interactions between features?

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

I was expecting that the loans with higher Debt to Income Ratio will have more 
instances of Bad Debt. But that did not seem to be the case. 

For unemployed people, mean credit score do not seem to be good indicator
their Loan Status. But for employed poeple, it seems to be good indicator.

The fact that loans with Income Range 'Not employed' and 'Not displayed' does not 
have any 'Current' loan status might mean that loan is no longer offered if the 
individual is unemployed.

------

# Final Plots and Summary


### Plot One
```{r echo=FALSE,  message=FALSE, warning=FALSE,Plot_One, fig.height=15,fig.width=8}


ggplot(aes(y=InquiriesLast6Months,x=CreditScoreBucket, 
           fill=LoanStatus_mod), 
       data = subset(loan,!(CreditScoreBucket %in% c("<500","900+")) &
                       !is.na(CreditScoreBucket))
      ) +
  geom_boxplot() +
  facet_grid(IncomeRange~.) +
  labs(x="Credit Score Range (Lower)",y="Mean Inquiries Last 6 Months", 
       title="Relationships of Credit Scores, Income Range, and\nInquiriesLast6Months to Loan Status"
       ) +
  coord_cartesian(ylim=c(0,10))

```

### Description One

This plot shows how Credit Scores, Income Range, and InquiriesLast6Months relates 
to Loan Status. Bad Loans have higher Number of Inquiries compared to the good ones,
especially at lower Credit Score ranges. As the Credit Score increases, the difference
becomes less and less obvious.

Completed loans have lower # of Inquiries than Bad Loans, but higher than Current
ones (especially for lower Credit Score range). This might indicate that investors 
for more recent loans are less likely to invest in loans with higher number of
inquiries and low credit scores.


------

### Plot Two
```{r echo=FALSE,  message=FALSE, warning=FALSE,Plot_Two, fig.height=4,fig.width=8}

ggplot(aes(x=IncomeRange,y=DebtToIncomeRatio,fill=LoanStatus_mod), data = loan) +
  geom_boxplot( ) +
  scale_colour_brewer(type = 'div')+
  coord_cartesian(ylim=c(0,1)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x="Income Range", y="Debt to Income Ratio", 
       title="Relationships of Income Range, Debt to Income Ration and Loan Status")


```

### Description Two
This plot shows that Debt to Income Ratio is not really a good indicator of Loan
Status. This shows that people with higher debt are not necessarily more likely to 
default on their loans.


------

### Plot Three
```{r echo=FALSE, message=FALSE, warning=FALSE,Plot_Three, fig.height=4,fig.width=8}


ggplot(aes(x=IncomeRange,y= CreditScoreRangeLower, fill=LoanStatus_mod  ), data = loan) +
  geom_boxplot(size=1) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + ylim(400, 800)+
  labs(x="Income Range", y="Mean Credit Scores",
       title="Relationships of Income Range, Credit Score and Loan Status")



```

### Description Three

This plot shows that Income Range and Credit Scores are good indicator of Loan Status,
but only for employed individuals. 

Individuals with lower income seem to have lower Credit Scores in general, and 
among them, the individuals with bad loans seem to have lower credit scores than 
the completed/current loans.

Surprisingly, unemployed people have higher credit scores overall than employed 
people. This might be because investors are only comfortable lending to unemployed
people if they have high credit scores. 

------

# Reflection

This exploration was primarily focused on identifying attributes that distinguished 
between Completed loan and Bad Loans. Current loans were also studied in some cases. 

Variables such as Prosper Score, Income Range, Credit Scores, and Inquiries made
in Last 6 Months were all related to Loan Status in varying degrees. There variables
are pretty good indicators to distinguish Bad loans from completed and current loans.
It is also seen that perhaps the investors of current loans are more risk-averse
than the investors for past completed & bad loans. Investors for current loans 
seem to be seeking out loans with higher credit scores and less number of
Inquiries in Past 6 Months.

The main struggle I had when performing the analysis was selecting the best 
plot to use to represent features. There are many varieties of plots available,
so it was lot of trial and error in determinng the best plot to convey information
about certain group of features.

The main success I had was in identifying the features that differentiated bad
loans from the good ones. I think I have a pretty good understanding of how 
different features are related to one another, and to the Loan Status. It was an
interesting exercise because I could pretend that I was one of the investors, and 
that I was trying to study which loans should I invest in, and which I should avoid.

However, more work needs to be done to identify strategies to generate the most
expected return for an investor. I believe valriables such as Borrower APR, Estimated
Return, Listing Category, time since loan origination, and other attributes about the
member can help investors maximize their returns.

